Accurate inference of microbial differential abundance from taxonomically-biased microbiome measurements

### 2021-09-09 abundance-measurement.Rmd
Move notes to temporary notes file

Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr) ^[todo: details about this assumption].
In this model, we suppose there are a set of *atomic taxa*, that differ in how efficiently they are measured but within which all cells behave similarly (r/v).
For concreteness, we equate atomic taxa with species. 
*(Note that bias may not be conserved w/in species; ref discussion of synthetic taxa later on)*.
We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.

Our model stipulates that the assigned read count of species $i$ in sample $a$ equal equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
\begin{align}
  (\#eq:measurement-model)
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.^[todo: add note about how, as defined, sequencing effort typically goes down with total density because of library normalization; also mention that it 'sequencing effort' is sometimes equated with total read count or library size, which is not what we've done here].
Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
\begin{align}
  (\#eq:total-reads)
  \text{total reads}(a)
    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
\end{align}
where 
\begin{align}
  (\#eq:mean-efficiency)
  \text{mean efficiency}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
\end{align}
is the average or mean efficiency of cells in the sample.

@@

The term *relative abundance* carries multiple meanings in microbiome science.
Microbiome researchers often use the relative abundance of a species in a sample to refer to its *proportion*, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g. Prokaryotes).
A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the *ratios* of cell densities among two or more species (e.g. a ratio of 5:4:2 among three species).
By this view, it is meaningless to refer to the relative abundance of an individual species.
<!-- (Although a proportion is also technically a ratio [of a species to a higher-level taxon], here we use 'ratio' to refer specifically to ratios among species.) -->
Although other interpretations exist ^[e.g. the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.

@@

- note: the error is not necessarily consistent for ratios among higher-order taxa
- note: The proportion equation is analogous to Equation \@ref(eq:ratio-error) (ratio case), with the denominator taxon is the entire taxonomic domain and the relevant efficiency is the weighted average of all species. In fact, both equations can be derived from a more general equation, where the observed ratio of two (potentially non-atomic) taxa equals the true ratio times the ratios of the mean efficiencies within each taxon.
- note: the results for ratios of higher-order taxa, and of proportions, can both be seen as stemming from a general result where the error in the ratio is the ratio of the mean efficiency within that taxon.
- perhaps have a box for this

@@

- note: right now I'm not consistently ref'ing the equations where the results come from

@@

- note: these two approaches are not always distinct, and it seems that many/most authors are thinking of spike-ins and housekeeping taxa in terms of the proportion-based method; I need to think about how to address this point.

>>>>

Our primary tool for understanding the impact of taxonomic bias MGS measurement is the theoretical model of MGS measurement developed and empirically validated by @mclaren2019cons.
This model is the simplest model of MGS measurement that includes multiplicative taxonomic bias while respecting the *compositional* nature of sequencing measurements, in which the total read count for a sample is unrelated to its total cell number or density (@gloor2017micr).
In this model, we suppose there are a set of *atomic taxa*, that differ in how efficiently they are measured but within which all cells behave similarly.
For concreteness, we equate atomic taxa with species. 
We ignore the possibility of contamination or false-positive taxonomic assignments---reads assigned to a given species and sample really do come from that species and sample.

Our model stipulates that the assigned read count of species $i$ in sample $a$ equal equals its cell density multiplied by a species-specific, sample-independent factor and a species-independent, sample-specific factor,
\begin{align}
  (\#eq:measurement-model)
  \text{reads}_{i}(a)
  = \text{density}_{i}(a) \quad \cdot
    \underbrace{\text{efficiency}_{i}}_{\substack{\text{species specific,} \\  \text{sample independent}}}
    \cdot \quad
    \underbrace{\text{sequencing effort}(a)}_{\substack{\text{species independent,} \\  \text{sample specific}}}.
\end{align}
The species-specific factor is the *relative measurement efficiency* (or *efficiency* for short) of the species, which represents how much more easily that species is measured (converted from cells to assigned reads) relative to an abitrarily chosen reference species (@mclaren2019cons).
The sample-specific factor, which we call the *sequencing effort* of the sample, reflects the fact that the number of reads per unit cell density varies among samples even in the absence of taxonomic bias due to variation in total density, library normalization, and total sequencing-run output.
Equation \@ref(eq:measurement-model) implies that the total reads from all species in the sample equal
\begin{align}
  (\#eq:total-reads)
  \text{total reads}(a)
    = \text{total density}(a) \cdot \text{mean efficiency}(a) \cdot \text{sequencing effort}(a),
\end{align}
where 
\begin{align}
  (\#eq:mean-efficiency)
  \text{mean efficiency}(a) 
  \equiv \frac{\sum_{j}\text{density}_j(a)\cdot \text{efficiency}_j}{\text{total density}(a)}
\end{align}
is the average or mean efficiency of cells in the sample.

@@

The term *relative abundance* carries multiple meanings in microbiome science.
Microbiome researchers often use the relative abundance of a species in a sample to refer to its *proportion*, or number of cells from that species divided by the total number of cells from the taxonomic domain of interest (e.g. Prokaryotes).
A second view, drawn from the statistical framework of Compositional Data Analysis, equates the relative abundances of species within a sample with the *ratios* of cell densities among two or more species (e.g. a ratio of 5:4:2 among three species).
By this view, it is meaningless to refer to the relative abundance of an individual species.
Although other interpretations exist^[e.g. the rank order within a sample, or value associated with a species after a centered-log-ratio transform], we focus on the proportion and ratio views as they underly most DA analysis methods (and microbiome analysis in general), for both relative and absolute abundances, and the differences between them help to clarify the issue over when and why the effects of bias cancel in an analysis.

@@

@@

@@

- Note: The current distinction between proportion- and ratio-based density estimation is incomplete, as the example of spike-in normalization shows: Whether the density of species $i$ is estimated by Equation \@ref(eq:density-ratio-est), or by first estimating total density from the ratio of non-spike-in to spike-in reads and then using Equation \@ref(eq:density-prop-est), the results are mathematically identical. Future revisions should clarify this point.

### 2021-09-09 real-world-implications.Rmd
Move notes to temporary notes file

- TODO: Address DA analysis of aggregate (or synthetic) taxa, such as phyla, which may be prone to errors due to changes in the species-level make-up of the taxon.

>>>>

- Todo: Address DA analysis of aggregate (or synthetic) taxa, such as phyla, which may be prone to errors due to changes in the species-level make-up of the taxon.

### 2021-09-09 revision-notes.Rmd
Move notes to temporary notes file

@@

The theoretical results from Section \@ref(differential-abundance) suggest a number of potential methods to avoid, correct, or otherwise mitigate the errors created by taxonomic bias.
We present five that seem particularly promising:
The first three are method for correcting or otherwise account for the effects of taxonomic bias to obtain more accurate DA results, whereas the second two treat the measurement efficiencies as unknown parameters and seek to understand the additional uncertainty they create, either in the results of individual studies or in joint analysis of multiple studies that used different MGS protocols.

@@

The most complete but experimentally challenging approach to mitigating taxonomic bias involves directly measuring the efficiencies of individual species using control samples (@mclaren2019cons).
Suitable controls are any community sample whose composition (species identity and relative abundances) has been previously characterized, either by construction (in the case of constructed 'mock' communities) or by measurement with a _reference protocol_ chosen to serve as a measurement standard.
These controls, when measured alongside the target experimental samples, can be used to estimate the relative efficiencies among the superset of species across the control samples.
The estimated efficiencies can then be used to calibrate the measured relative abundances among the control species in the target samples to the chosen measurement standard, by simply dividing the species' read counts by their corresponding efficiencies or using a more sophisticated statistical modeling framework.
These calibrated compositions can then be used for arbitrary downstream microbiome analyses, including DA analysis.
Calibration of species not in the controls requires first imputing their efficiencies from the control measurements, for example using phylogenetic relationships, genetic characteristics (such as 16S copy number), and/or phenotypic properties (such as cell-wall structure).
Though such imputation is conceptually straightforward to implement, its reliability and utility remain untested.

@@

The DA analyses we consider don't need accurate estimates of relative abundances within individual samples---they only require accurate fold changes across samples.
Section \@ref(differential-abundance) showed that fold changes in species ratios and ratio-based density estimates can remain accurate even when taxonomic bias distorts the values in individual samples.
Hence a solution is to use to analyze variation in ratios and ratio-based density estimates.
For the latter, practical options include

- spike-ins of extraneous species at constant and/or known density
- housekeeping taxa, stipulated or computationally identified as having a constant density across samples
- targeted measurement of specific taxa

TBD what further discussion or illustration to have here.

### notes

@@

In some cases, we may still which to use a proportion-based approach to analyzing variation in density.
For example, in some systems measuring total cell concentration may be easier than implementing a spike-in experiment or determining a set of reference species to directly measure for which at least one species is present in all samples.
In this case, targeted measurements of one or more reference species can be used to validate and/or correct the fold changes or regression coefficients infered from the proportion-based densities.
Recall that the error in proportion-based log fold changes (LFCs) is shared among species and is equal to the inverse LFC fold change in mean efficiency (Equations \@ref(eq:prop-fc-error) and \@ref(eq:regression-error); Figure \@ref(fig:regression-example)D).
Consequently, an accurate LFC of one or more reference species can serve as a general validation or used to correct the LFCs of all species.

- NOTE: I'm not sure if I really understand this example below, since I don't see how this interpretation of the Table 1 results mesh with Figure 4

We illustrate this idea with an example from the study by @lloyd2020evid described in Section \@ref(implications), which estimated doubling times for various taxa from the measured log increase in cell density with burial time in marine sediments.
The primary analysis, which used proportion-based density estimates, was supported by targeted measurements (qPCR) of 16S concentration for several bacterial and archeal taxa (CHECK).
In Core 30, the doubling time estimated for the Archaeal phylum Bathyarchaeota was estimated to be 10.1 years (CHECK) using the proportion-based densities and 10.3 years by qPCR (@lloyd2020evid Table 1 and Figure 4).
These doubling times correspond to highly similar estimated slopes of log density per year of $1/10.1 \approx 0.99$ and $1/10.3\approx 0.97$.
The difference between these slopes, $\approx 0.02$, estimates the slope of the mean efficiency.
It's small value suggests little systematic variation in mean efficiency with burial time and hence little systematic bias in the estimated doubling times derived from the proportion-based density estimates.
In this case, close agreement in the DA results across methods for a specific taxon suggests that variation in sample mean efficiency did not significantly distort the MGS-based results and hence provides a degree of validation for all taxa, not just the taxa with targeted measurements.
(An important caveat for this example is that this analysis assumes consistent efficiencies at the higher taxonomic levels used for qPCR measurement and MGS analysis performed by @lloyd2020evid.)
Here we compared summary statistics; however, in principle a joint statistical analysis of targeted and proportion-based density measurements would provide more statistical power to estimate and correct the effects of variation in the mean efficiency across samples and provide calibrated fold change and regression estimates for all taxa [APPENDIX].

@@

What can we do for experiments that have been conducted without control communities, targeted control measurements, or spike-ins?
One approach is to conduct a bias sensitivity analysis to probe the extent and way in which the results of a DA analysis are sensitive to different possible taxonomic biases in our experiment.

A straightforward approach to bias sensitivity analysis is to use computer simulation to randomly generate efficiency vectors, each of which is used to calibrate the MGS measurements.
These calibrated MGS measurements represent the true compositions under the hypothesized taxonomic bias. 
By re-running our DA analysis and comparing results across these simulated datasets, we can begin to understand how our initial results might be influenced by the strength and structure of the taxonomic bias in our experiment.

The simulated efficiency vectors may be generated to reflect certain hypotheses about bias in the given system.
The utility of such simulations can increase the more we learn about the magnitude of bias in different systems and the taxonomic and protocol features that determine it.
Future work in developing tools and methods for simulating efficiency vectors and performing bias sensitivity analyses could be a valuable way to assay and improve the reliability of microbiome results—for differential abundance and microbiome analyses more generally.

@@

So far we have considered analysis of samples subject to the same taxonomic bias;
however, meta-analysis of microbiome samples measured across multiple studies must often contend with a diversity of protocols and hence taxonomic biases in different groups of samples.
This inconsistency in taxonomic bias across studies poses a problem even for ratio-based analyses where the effects of bias might cancel within a study.

A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects" --- amorphous differences in the measurements of each study.
By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.

### notes

>>>>

# Notes from September 2021 revision

# How taxonomic bias affects abundance measurements {#abundance-measurement}

- Footnote: "This model is the simplest model...total cell number or density (@gloor2017micr) ^[todo: details about this assumption]."
- Footnote: "The sample-specific factor...and total sequencing-run output.^[todo: add note about how, as defined, sequencing effort typically goes down with total density because of library normalization; also mention that it 'sequencing effort' is sometimes equated with total read count or library size, which is not what we've done here]."
- on assumption of conserved efficinecy within speies: Note that bias may not be conserved w/in species; might add and ref discussion of synthetic taxa later on

## Relative abundances (proportions and ratios)

- after defining species ratios, might note: (Although a proportion is also technically a ratio [of a species to a higher-level taxon], here we use 'ratio' to refer specifically to ratios among species.)

- note: the error is not necessarily consistent for ratios among higher-order taxa
- note: The proportion equation is analogous to Equation \@ref(eq:ratio-error) (ratio case), with the denominator taxon is the entire taxonomic domain and the relevant efficiency is the weighted average of all species. In fact, both equations can be derived from a more general equation, where the observed ratio of two (potentially non-atomic) taxa equals the true ratio times the ratios of the mean efficiencies within each taxon.
- note: the results for ratios of higher-order taxa, and of proportions, can both be seen as stemming from a general result where the error in the ratio is the ratio of the mean efficiency within that taxon.
- perhaps have a box for this

## Absolute abundances (cell densities)

- note: right now I'm not consistently ref'ing the equations where the results come from
- note about ratio-based density inference: see note left in the abundance-measurement.Rmd

# How taxonomic bias affects differential-abundance analysis {#differential-abundance}

- Maybe: Add a final sentence summary that the constant multiplicative error in ratio-based abundance estimates does not affect DA measurement, but the varying error induced by variation in the mean efficiency affects DA analysis on proportion-based estimates.

## Change between a pair of samples

## Regression analysis of many samples

- Important complications to mention:
  - Error in total density or targeted density estimates
  - Counting error (so far ignored) can lead to missed associations of species with low measurement efficiencies. Example: G. vaginalis associations are not detected by several studies that use certain V13 primers that simply don't amplify this species well. This is a case where the precision in estimated associations is dramatically reduced, by a different mechanism than what is described above.

@@

@@

@@

@@

- Note: I may not correctly understand the next example, since I don't yet see how this interpretation of Table 1 meshes with Figure 4

@@

@@

### 2021-09-09 solutions.Rmd
Move notes to temporary notes file

- Somewhere note that we're considering ideas beyond 'MGS protocol optimization', but that is still important - tuning protocols to reduce taxonomic bias, and to make it as consistent as possible across samples. (Our results confirm the importance of both of these)

@@

- The above states the gist; can also have a paragraph with description/discussion of how this is/has been used so far, and ongoing developments and challenges. Some bullets for this follow.
- A mock-community approach may be sufficient for synthetic-community experiments, where all taxa are culturable, and for relatively simple natural communities like the vaginal microbiome that are dominated by a small number of culturable taxa.
- But suitable mock controls may not be feasible for most complex natural ecosystems and require significant effort to develop.
- An closely-related alternative to mocks are controls derived from natural samples, which provide a way to calibrate measurements from protocols to a reference protocol (@mclaren2019cons).
- A natural fecal standard is currently being developed by [NIST](https://www.nist.gov/programs-projects/human-gut-microbiome-reference-material) and one has recently been made available commercially by Zymo Research ([ZymoBIOMICS Fecal Reference](https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards/products/zymobiomics-fecal-reference-with-trumatrix-technology)).
- Ensuring accurate quantification, stability, homogeneity of such controls remains a major challenge.
- Careful testing is also needed to ensure that the preparation and storage of the controls has not significantly affected the taxonomic bias relative to the experimental samples in a given application.
- As it is feasible to characterize a single standard much more extensively than a typical community sample, we may be able to obtain an estimated composition we feel comfortable treating as the ground truth. 
- Yet even when this is not possible, such natural standards can allow us to reconcile results across studies despite not knowing the truth.
- Given the significant uncertainty about estimated efficiencies (stemming from all the above raised caveats), rather than as providing calibration to the 'truth', the estimated efficiencies might be best seen as priors or hypotheses to inform a bias-sensitivity analysis of the sort described below.
- Should mention the two extant cases where mocks of relevant taxa have been created: The vaginal-HMP / MOMS-PI study (@brooks2015thet and @fettweis2019thev) and the @leopold2020host plant-fungal experiment. Note that even in these cases, calibration hasn't been much used, partly because of a lack of workflows or understanding of when and why it is necessary. Ideally we can give an illustration; this section can be a good use of the @leopold2020host change-over-time result, though I'm not sure whether to discuss that here or in the 'real-world' section.

@@

### notes

For spike-ins, a crucial consideration may be whether the spike-in is added pre- or post-DNA extraction (ref appendix, Harrison).
Similarly, we must consider whether targeted measurements are made pre extraction (e.g. CFU counting or ddPCR directly on cells) or post extraction (e.g. ddPCR or qPCR of DNA).
If DNA yield is linear, then either approach should give reliable fold changes; but if DNA yield is non-linear (for example, saturating in higher biomass samples) then systematic errors can still arise with the post-extraction approach, though these may still be reduced compared to proportion-based inference (REF appendix).
(Why expect to be reduced: This saturation will still be less than what would be imposed by library normalization, though I need to think about this a bit.)
In additional, substantial random variation in the (log) DNA yield of a sample with a given biomass will reduce the precision of the estimated abundances that is possible with the post-extraction approach in a manner that cannot be circumvented by increasing the number of spike-in or targeted species.

Notes on implementation:

- Targeted measurement 
- Although solutions can be hacked together, ideally we'd have new statistical tools to jointly model MGS and supplemental measurements or spike-ins and housekeeping taxa.
- These would account for the uncertainty associated with the targeted measurements, including detection limits, and could naturally allow for handling the fact that targeted taxa may not be present in every sample.
- The appendix gives some theoretical insight into how to deal with a given reference taxon not being present or detectable in every sample

@@

- NOTE: I'm not sure if I really understand this example below, since I don't see how this interpretation of the Table 1 results mesh with Figure 4

@@

- Aim to illustrate with a simulated or real example

@@

A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects" --- amorphous differences in the measurements of each study.
By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.

### notes

When the goal is to perform a meta-analysis that combines studies that have used different protocols, the unknown measurement efficiencies of each protocol can be explicitly included as parameters of the statistical model that is used.
Unknown efficiencies can be included in "compositional" linear modeling frameworks such as ALDEx2, DivNet, and fido simply by adding a protocol-specific term to the linear model of taxon log ratios.
[*Add sentence on how this can be used for non-compositional analyses.*]
Thus such bias-aware meta-analyses are already technically feasible and—if bias is truly consistent within a protocol or study—may provide a more powerful alternative to non-parametric or other meta-analysis methods that do not model bias explicitly.

- Aim to  illustrate with a simulated or real example

- Value: If results of individual studies might be fairly robust to unknown bias, then might be able to efficiently (from a statistical POV) pool studies when they biologically agree or rule out bias (under the used assumptions) as the cause of disagreement when they don't agree.

>>>>

@@

- The above states the approach; can add description/discussion of application to the MOMS-PI and @leopold2020host cases and developments and challenges in availability of control communities

@@

@@

- Note: I may not correctly understand the next example, since I don't yet see how this interpretation of Table 1 meshes with Figure 4

@@

@@

A potential solution is to perform a bias-aware meta-analyses that jointly analyzes multiple data from studies while explicitely acknowledges the distinct, unknown taxonomic bias in each study.
Parametric microbiome meta-analysis models already include sets of species-specific latent parameters to model "batch effects"---amorphous differences in the measurements of each study.
By configuring these models so that these latent parameters correspond to the study and species specific efficiencies, we can increase the biological interpretability of these models and gain ability to determine whether disageements in DA results across studies are due to differences in taxonomic bias or other differences (such as host cohort or study design).
Using a biophysically motivated model parameterization may also lead to performance gains in terms of increased explanatory power for a given model complexity, as seen for models of mock community measurements by @mclaren2019cons, which can increase improve our ability to find reproducible microbial associations in a given meta-study.
